Implementing simplified INLA into `aghq`

### 2022-10-10 src/dev_sinla/sinla.Rmd
Getting there! Still some issues to fix r.e. normalising and weights, cleaning up

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
$$
where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by exploration of $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
where $H_k(z_d)$ are the Hermite polynomials for $k \in \mathbb{N}$, and
$$
w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ with nodes $z$, of which there will be $k^d$.
Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triange of the curvature at a randomly chosen quadrature point).
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.

@@

```{r}
mcmc <- tmbstan::tmbstan(h, chains = 4, iter = 5000)
```

@@

```{r}
extract(mcmc, pars = "phi_prev[1]") %>%
  as.data.frame() %>%
  ggplot(aes(x = phi_prev.1.)) +
    geom_histogram(aes(y = ..count.. / sum(..count..)), alpha = 0.8, fill = cbpalette[7]) +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior")
```

@@

## $p(\theta \, | \, y)$

@@

## Moving to $\tilde \tilde p(x_i \, | \, y)$

Let's have a go at computing $\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z)$ so that we can compare to the posterior distribution from `tmbstan`.

We have simplified matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{G}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{G}(x_i \, | \, \hat \theta, y)
\end{align}
$$
We're going to evaluate this by taking samples from the full Gaussian, the subsetting them:

```{r}
#' To write
```

@@

@@

  #' random are integrated out with a Laplace approximation
  obj <- MakeADFun(
    data = dat_i,
    parameters = param_i,
    random = "phi_prev_minus_i",
    DLL = "model1index"
  )

@@

  sd_out <- sdreport(
    obj,
    par.fixed = opt$par,
    getJointPrecision = TRUE
  )

  #' Parameters of the Gaussian approximation
  mean <- with(obj$env, last.par[random])
  Q <- obj$env$spHess(with(obj$env, last.par), random = TRUE)

@@

We should be able to reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
For example:

@@

@@

$$
\hat \mu_{-i}(x_i, \theta) = [\hat \mu(\theta)]_{-i} + \hat Q(\theta)^{-1}_{-i, i} \hat Q(\theta)_{i, i}(x_i - [\hat \mu(\theta)]_{i})
$$

>>>>

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by exploration of $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
where $H_k(z_d)$ are the Hermite polynomials for $k \in \mathbb{N}$, and
$$
w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ with nodes $z$, of which there will be $k^d$.
Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\begin{equation}
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxigiveny)
\end{equation}
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.

@@

```{r}
mcmc <- tmbstan::tmbstan(h, chains = 4, iter = 5000, refresh = 0)
```

@@

```{r}
plot <- extract(mcmc, pars = "phi_prev[1]") %>%
  as.data.frame() %>%
  ggplot(aes(x = phi_prev.1.)) +
    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior")

plot
```

@@

## $\tilde p(\theta \, | \, y)$

@@

## $\tilde p(x_i \, | \, y)$

Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{G}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{G}(x_i \, | \, \hat \theta, y)
\end{align}
$$
We can evaluate this for $i = 1$ via:

```{r}
i <- 1
mean_i <- as.numeric(mean[i])
var_i <- solve(Q)[ i, i]
```

And compare it to our MCMC results as follows:

```{r}
plot2 <- plot +
  stat_function(
    data = data.frame(x = c(0.5, 3.5)), aes(x), fun = dnorm, n = 101, 
    args = list(mean = mean_i, sd = sqrt(var_i)), col = cbpalette[2], size = 1) +
  labs(x = "prev_phi[1]", y = "Posterior") +
  theme_minimal()

plot2
```

@@

## $\tilde p(x_i \, | \, \theta, y)$

@@

  #' random are integrated out with a Laplace approximation
  obj <- MakeADFun(
    data = dat_i,
    parameters = param_i,
    random = "phi_prev_minus_i",
    DLL = "model1index",
    silent = TRUE
  )

@@

  #' Parameters of the Gaussian approximation
  mm <- obj$env$last.par
  mean <- mm[obj$env$random]
  Q <- obj$env$spHess(mm, random = TRUE)

@@

# Reproducing output from `obj`

As above, we should be able to reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
For example:

@@

## $\tilde p(x_i \, | \, y)$

Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{LA}(x_i \, | \, \hat \theta, y)
\end{align}
$$

Start by getting $\hat \theta$:

```{r}
opt <- nlminb(
  start = h$par,
  objective = h$fn,
  gradient = h$gr,
  control = list(iter.max = its, trace = 0)
)

#' How many equally spaced points do we want to evaluate the posterior at?
phi_prev_i_grid_length <- 50

df <- opt$par %>%
  t() %>%
  as.data.frame() %>%
  slice(rep(1:n(), each = phi_prev_i_grid_length)) %>%
  mutate(
    phi_prev_i = seq(from = 0.5, to = 3.5, length.out = phi_prev_i_grid_length)
  ) %>%
  rowwise() %>%
  mutate(
    #' Note here: it's crucial that the arguments are in the right order!
    logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev))
  )

ggplot(df, aes(x = phi_prev_i, y = logpost)) +
  geom_line() +
  theme_minimal() +
  labs(x = "phi_prev[1]", "log(posterior)")
```

What we've got here is the unnormalised log-posterior?
How can we go from that to a normalised posterior?
Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.

$$
\begin{align}
\tilde f(x) &= \frac{f(x)}{\int f(x) \text{d}x} = \frac{f(x)}{C}
\implies \\
\log\tilde f(x) &= \log f(x) - \log C
\end{align}
$$

```{r}
df <- df %>%
  mutate(
    logpost_mean = colMeans(select(., logpost)),
    logpost_adj = logpost - logpost_mean,
    post = exp(logpost_adj)
  )

plot3 <- plot2 +
  geom_line(data = df, aes(x = phi_prev_i, y = post), col = cbpalette[5], size = 1)

#' Need to normalise this!
plot3
```

@@

This remains to-do!
